manuscript

### 2023-06-01 manuscript.rmd
added to all sections

**\[Introduce sound, soundscapes, noise in the ocean, noise pollution\]**\
@pijanowski2011 (+ mapping soundscapes). Soundscapes are made up of three broad categories of sound sources: \
- biophonies, which come from marine mammals, fishes, and invertebrates; \
- geophonies, which consist of natural sounds like rain, wind, earthquakes; \
- anthropophonies, which consist of human-made sounds from vessels, offshore development, active sonar pingers, and seismic surveys (@miksis-olds, @duarte2021).

@@

\[Chronic/residual/ambient/transient/continuous/impulsive sound\] @mckenna2021 and @richardson2013 \
Transient sounds have a short duration, while continuous sounds keep going. @richardson2013.

Lower frequencies travel further, and higher frequencies tend to attentuate (@richardson2013 ), so the focus should be on local effects for that (@hildebrand2009 ). Ambient noise in shallow water has more variability than deeper water (@richardson2013)

@@

**\[Introduce beaked and sperm whales: protected, visually cryptic, animals that are sensitive to noise and use sound extensively to communicate\]**\
[Beaked whales]{.underline} (family) routinely dive for hours at a time to depths greater than 1000 ft. They can dive for over 3 hours, followed by only a few minutes on the surface (same for BW and SW?) ( see Watkins et al., 1993). This diving behavior, coupled with a low-profile body makes them nearly invisible on the surface, so very little is known about them. This may be an anti-predation adaptation to avoid killer whales (cite) as well as a synchronized foraging dives following a silent descent (@aguilardesoto2020 ). Because they must make sound to locate prey, they do so in the refuge of deep depths where killer whales do not go (less than 20 m @aguilardesoto2020 and in frequencies they do not hear (cite). It is thought that this anti-predator response is exactly what makes them sensitive to human noise.

[Sperm whales]{.underline} (*Physeter macrocephalus*) may have more acoustic tolerance than beaked whales (Miller et al. 2022).

The acoustic interference from vessels, for example, can :\
- affect behavior (displacement from habitat, spending time and effort fleeing instead of foraging (Joy et al. 2019, @deruiter2013, @aguilarsoto)\
- disrupt communication and echolocation\
- lead to physical harm, even at far distances (See @hooker2019 for more info) like hearing loss and elevated stress levels\
- masking signals for communication (\@erbe, or environmental cues (@duarte2021)\
- reduced mating calls (Weilgart, 2007)

@@

Despite their visually cryptic nature, their sound signals are very well-described, down to the species level @baumann-pickering2013, @baumann-pickering2014. Beaked whale produce clicks, pulses, and buzzes during foraging starting at the 400(?) m range @tyack. The production of sound at this depth may be an adaptation to avoid predation from killer whales, who spend most of its time shallow (from @tyack Baird et al 1998, follow up on this). Some are still being discovered and identified - see @manzano-roth2023.

@@

**\[Introduce the importance of passive acoustic monitoring on this type of study, gaps in visual survey methods\]**\
Passive acoustic monitoring is a tool that allows us to record and listen to the soundscape. Beaked and sperm whales are ideal study candidates for this type of monitoring because they are often better heard than seen. **\[List pros and cons of all types of PAM: acoustic tags, towed hydrophones, bottom-mounted, static (Mellinger et al)\].** One type of passive acoustic recorder called a drifting buoy allows us to capture the dynamic nature of sound over a large spatial area and especially in the deep, offshore, and less studied waters. Studying beaked whales can be challenging due to how offshore they live (@robbins2022). It can drift freely with the currents for days or weeks at a time. The goal with passive acoustic monitoring is to improve current species distribution models of beaked and sperm whales that utilize visual survey methods only. Historically, it has been challenging to record foraging calls at several hundred meters of depth @barlow2005.

@@

Automatic Identification System. This was collected to investigate patterns of co-occurrence between vessels and whales and sound levels in those moments.

@@

@@

**\[Goal\]**\
Characterize the acoustic habitat of beaked and sperm whales in the California Current Ecosystem using soundscape metrics and vessel tracking. Characterizing the acoustic environment and identifying co-occurrence of whales and anthropogenic sound has been done in a similar study by @weiss2021.

@@

Both of these outcomes are of high importance to managers and agencies especially NOAA. Within their Ocean Noise Strategy Roadmap, NOAA is seeking 1) a quantification of the spatial and temporal variability of ambient noise conditions, and 2) understanding of how anthropogenic soudn sources contribute to the soundscape. Co-occurrence of buoys and vessels will provide fine-scale snapshots of sound levels surrounding vessels and concurrent whale foraging activity. With AIS, we can calculate distance from buoy to vessel and sound levels in those moments. And 3) NOAA is seeking more information on vocally active species' distributions.

@@

**\[Overall data collection efforts\]**\
Acoustic detections were collected with the use of free-drifting Drifting Acoustic Spar Buoy Recorders (hereafter buoys). These were deployed as part of the California Current Ecosystem Survey (CCES) spanning the offshore waters from southern Canada to northern Mexico in summer of 2018. Buoys offer greater spatial and temporal coverage compared to bottom-mounted or stationary hydrophones, ideal for detection of offshore or mobile species. These buoys contained two satellite geo-location trackers, two hydrophones, etc and collected a total of 1910 hours of data. SWFSC ran acoustic detections in PAMGuard. Eight buoys were lost at sea and therefore removed from the analysis. These contain 1, 5, 10, 25, 75, 95, and 99 percentile distribution from each metric.

@@

**\[Post-processing details, data cleaning\]**\
Soundscape metrics were rounded to the nearest 20 minutes (:00, :20, :40) and converted to UTC. These include third-octave level bands and broadband (20-24,000 Hz) metrics. ~~One way to measure sound is with a metric called a third octave level, which splits the sound spectrum into many bands.~~ The band at 125 Hertz is often associated with shipping vessel noise and the 2000 Hertz or 2 kHz with mid-frequency naval sonar (cite). Beaked whales are known to be sensitive to sonar in this latter range (cite). Several TOL levels were chosen as proxies to vessel or sonar frequencies (TOL 63 to 20,000).

Whale detections were rounded to nearest 20 minutes. A separate column was created as a common join field that contained nearest timestamp in UTC to that of drifter tracks. Two types of BW species (?BW and ?BWC) were removed from analysis as the species is unknown. Whale detections were reduced to "presence/absence" from number of clicks, estimated individuals, etc.

@@

| Covariate                     | Unit | Source                                                          | Justification |
|--------------|--------------|-------------------------------|--------------|
|                               |      | \[can use A, B, C to avoid repetition, cite in caption\]        |               |
| Distance to continental slope |      | General Bathymetric Chart of the Oceans (GEBCO) @weatherall2015 |               |
| Bathymetric slope             |      | GEBCO tif                                                       |               |
|                               |      |                                                                 |               |
|                               |      |                                                                 |               |
|                               |      |                                                                 |               |
|                               |      |                                                                 |               |
|                               |      |                                                                 |               |
|                               |      |                                                                 |               |
|                               |      |                                                                 |               |
|                               |      |                                                                 |               |
|                               |      |                                                                 |               |
|                               |      |                                                                 |               |
|                               |      |                                                                 |               |

@@

\[Describe simple statistical relationships, correlations, exploratory data findings (highest, lowest, etc., time of day, drift with most detections/day)\]

@@

>>>>

**\[Sound, soundscapes, noise in the ocean, noise pollution\]**\
@pijanowski2011 (+ mapping soundscapes). Soundscapes are made up of three broad categories of sound sources:\
- biophonies, which come from marine mammals, fishes, and invertebrates;\
- geophonies, which consist of natural sounds like rain, wind, earthquakes;\
- anthropophonies, which consist of human-made sounds from vessels, offshore development, active sonar pingers/echosounders, and seismic surveys (@miksis-olds, @duarte2021).

@@

\[Chronic/residual/ambient/transient/continuous/impulsive sound\] @mckenna2021 and @richardson2013\
Transient sounds have a short duration, while continuous sounds keep going. @richardson2013.

Lower frequencies travel further, and higher frequencies tend to attenuate (@richardson2013 ), so the focus should be on local effects for that (@hildebrand2009 ). Ambient noise in shallow water has more variability than deeper water (@richardson2013)

@@

**\[Introduce beaked and sperm whales: protected, visually cryptic, animals that are sensitive to noise and use sound extensively to communicate\]**\
[Beaked whales]{.underline} (*Ziphiidae*) routinely dive for hours at a time to depths greater than 1000 ft. They can dive for over 3 hours, followed by only a few minutes on the surface (same for BW and SW?) ( see Watkins et al., 1993). This diving behavior, coupled with a low-profile body makes them nearly invisible on the surface, so very little is known about them. This may be an anti-predation adaptation to avoid killer whales (cite) as well as a synchronized foraging dives following a silent descent (@aguilardesoto2020 ). Because they must make sound to locate prey, they do so in the refuge of deep depths where killer whales do not go (less than 20 m @aguilardesoto2020 and in frequencies they do not hear (cite). Why do they have both adaptations? It is thought that this anti-predator response is exactly what makes them sensitive to human noise.

Species studied:\
- ZC Cuvier's beaked whale *Ziphius cavirostris\
-* BW43, possibly Perrin's beaked whale *M*. *perrini*\
- BW37V, possibly Hubbs' beaked whale *M. carlhubbsi\
-* BB, --???\
- BWC, possibly ginkgo-toothed beaked whale *M. ginkgodens\
-* PM, sperm whale

sp[erm whales]{.underline} (*Physeter macrocephalus*) may have more acoustic tolerance than beaked whales (Miller et al. 2022).

The acoustic interference from human activity (IN ODONTOCETES OR ALL MARINE LIFE?), for example, can :\
- affect **behavior** (displacement from habitat, spending time and effort fleeing instead of foraging (Joy et al. 2019, @deruiter2013, @aguilarsoto2006 , @stanistreet2022, @joyce2020, @falcone2017, @cholewiak2017 )\
- disrupt **communication** and echolocation (same as above?)\
- lead to physical **harm**, even at far distances (up to 100 km away from @falcone2017) like hearing loss, tissue damage, and elevated stress levels. (See @hooker2019 for more info)\
- **masking** signals for communication (@erbe2014?, ) or environmental cues (@duarte2021)\
- reduced mating calls (Weilgart, 2007)\
- see Cox et al 2006 for impacts of sound on BW

@@

Despite their visually cryptic nature, their sound signals are very well-described, down to the species level @baumann-pickering2013, @baumann-pickering2014. Beaked whale produce clicks, pulses, and buzzes during foraging starting at the 400(?) m range @tyack. The production of sound at this depth may be an adaptation to avoid predation from killer whales, who spend most of its time shallow (from @tyack Baird et al 1998, follow up on this). Some are still being discovered and identified globally- see @manzano-roth2023, @yamada2019.

Do beaked whales show temporal variation in their calls? See @ziegenhorn2023

@@

**\[Introduce the importance of passive acoustic monitoring on this type of study, gaps in visual survey methods\]**\
Passive acoustic monitoring is a tool that allows us to record and listen to the soundscape. Beaked and sperm whales are ideal study candidates for this type of monitoring because they are often better heard than seen.

**\[List pros and cons of all types of PAM: acoustic tags, towed hydrophones, bottom-mounted, static (Mellinger et al)\].** One type of passive acoustic recorder called a drifting buoy that drifts freely with the currents for days or weeks at a time. It allows us to capture the dynamic nature of sound over a large spatial area and especially in the deep, offshore, and less studied waters. Buoys offer greater spatial and temporal coverage compared to bottom-mounted or stationary hydrophones, ideal for detection of offshore or mobile species. Studying beaked whales can be challenging due to how offshore they live (@robbins2022). The goal with passive acoustic monitoring is to improve current species distribution models of beaked and sperm whales that utilize visual survey methods only. Historically, it has been challenging to record foraging calls at several hundred meters of depth @barlow2005.

@@

Automatic Identification System. Ships are abundant throughout the study area due to major US ports. This was collected to investigate patterns of co-occurrence between vessels and whales and sound levels in those moments.

@@

How have other studies separated vessel noise (contribution to soundscape) from ambient soundscape? See @weiss2021 and @haver2020. Power spectral density ambient sound levels.

@@

**\[Goal\]**\
Characterize the acoustic habitat of beaked and sperm whales in the California Current Ecosystem using soundscape metrics and vessel tracking. Characterizing the acoustic environment and identifying co-occurrence of whales and anthropogenic sound has been done in a similar study by @weiss2021. This serves as baseline monitoring of soundscape for offshore areas.

@@

Both of these outcomes are of high importance to managers and agencies especially NOAA. Within their Ocean Noise Strategy Roadmap, NOAA is specifically seeking 1) a quantification of the spatial and temporal variability of ambient noise conditions, and 2) understanding of how anthropogenic soudn sources contribute to the soundscape. Co-occurrence of buoys and vessels will provide fine-scale snapshots of sound levels surrounding vessels and concurrent whale foraging activity. With AIS, we can calculate distance from buoy to vessel and sound levels in those moments. And 3) NOAA is seeking more information on vocally active species' distributions.

@@

**\[Overall data collection efforts\]**\
Acoustic detections were collected with the use of free-drifting Drifting Acoustic Spar Buoy Recorders (hereafter buoys). These were deployed as part of the California Current Ecosystem Survey (CCES) spanning the offshore waters from southern Canada to northern Mexico in summer of 2018. These buoys contained two satellite geo-location trackers, two hydrophones, etc and collected a total of 1910 hours of data. SWFSC ran acoustic detections in PAMGuard. Eight buoys were lost at sea and therefore removed from the analysis. These contain 1, 5, 10, 25, 75, 95, and 99 percentile distribution from each metric.

@@

**\[Post-processing details, data cleaning\]**\
Soundscape metrics were rounded to the nearest 20 minutes (:00, :20, :40) and converted to UTC. These include third-octave level bands and broadband (20-24,000 Hz) metrics. ~~One way to measure sound is with a metric called a third octave level, which splits the sound spectrum into many bands.~~ The band at 125 Hertz is often associated with shipping vessel noise and the 2-5 kHz range with mid-frequency naval sonar ( @garrett2016 ). Beaked whales are known to be sensitive to sonar in this latter range (cite). Several TOL levels were chosen as proxies to vessel or sonar frequencies (TOL 63 to 20,000).

Whale detections were rounded to nearest 20 minutes. A separate column was created as a common join field that contained nearest timestamp in UTC to that of drifter tracks. Two types of BW species (?BW and ?BWC) were removed from analysis as the species is unknown and unsure. Whale detections were reduced to "presence/absence" from number of clicks, estimated individuals, etc.

@@

| Covariate                     | Unit | Source                                                          | Justification |
|-----------------|-----------------|----------------------|-----------------|
|                               |      | \[can use A, B, C to avoid repetition, cite in caption\]        |               |
| Distance to continental slope |      | General Bathymetric Chart of the Oceans (GEBCO) @weatherall2015 |               |
| Bathymetric slope             |      | GEBCO tif                                                       |               |
|                               |      |                                                                 |               |
|                               |      |                                                                 |               |
|                               |      |                                                                 |               |
|                               |      |                                                                 |               |
|                               |      |                                                                 |               |
|                               |      |                                                                 |               |
|                               |      |                                                                 |               |
|                               |      |                                                                 |               |
|                               |      |                                                                 |               |
|                               |      |                                                                 |               |
|                               |      |                                                                 |               |

@@

[Describe simple statistical relationships, correlations, exploratory data findings (highest, lowest, etc., time of day ( @ziegenhorn2023 ) , drift with most detections/day)]

@@

Look at power spectral density? @haver2020
